---
title: "Problem Set 1"
author: "Yipeng Liu"
format: html
editor: visual
---

# Homework 1

github page: https://github.com/EfraimLYP/stats506

## Problem 1

### Step 1

First, convert the .data file into the data frame form and give appropriate column names.

```{r}
# Import the data from the .data file
wineDf <- read.table("/Users/yipengliu/Desktop/Graduate Courses/STATS 506/Homework/Problem Set 1/wine/wine.data", sep = ",")
colnames(wineDf) <- c("Type", 
                      "Alcohol", 
                      "MalicAcid", 
                      "Ash",
                      "AlcalinityOfAsh",
                      "Magnesium",
                      "TotalPhenols",
                      "Flavanoids",
                      "NonflavanoidPhenols",
                      "Proanthocyanins",
                      "ColorIntensity",
                      "Hue",
                      "OD280OD315OfDilutedWines",
                      "Proline")

# Display the first 5 rows of the data frame
head(wineDf, 5)
```

### Step 2

Check the wine with the highest and lowest alcohol content belongs.

```{r}
# Type of wine contains the highest alcohol content
typeContainsMostAlcohol <- 
  wineDf$Type[which(wineDf$Alcohol == max(wineDf$Alcohol))]
typeContainsMostAlcohol

# Type of wine contains the lowest alcohol content
typeContainsLowestAlcohol <- 
  wineDf$Type[which(wineDf$Alcohol == min(wineDf$Alcohol))]
typeContainsLowestAlcohol
```

It follows that the wine with the highest alcohol content belongs to class 1, and that the wine with the lowest alcohol content also belongs to class 1.

Then, explore how many wines in our example have higher levels of magnesium than average German beers.

```{r}
quantityHigherThanGerman <- sum(wineDf$Magnesium > 114)
quantityHigherThanGerman
```

It follows that there are 26 wines have higher levels of magnesium than German beers.

Finally, find out within each class, how many wines have higher levels of magnesium than average German beers.

```{r}
# Create a new column for the categorial variable "whether of not it has higher levels of magnesium than German beers"
wineDf$HigherThanGermanBeers <- 
  as.numeric(wineDf$Magnesium > 114)

# Calculate within each class, how many wines have higher levels of magnesium than average German beers
sumHigherThanGermanBeers <- 
  tapply(wineDf$HigherThanGermanBeers, 
         INDEX = wineDf$Type, 
         FUN = sum)
sumHigherThanGermanBeers
```

It can be seen that 15, 6, and 5 wines in categories 1, 2, and 3, respectively, have higher magnesium contents than average German beers.

### Step 3

Here creates a table identifying the average value of each variable, providing one row for the overall average, and one row per class with class averages.

```{r}
# Get the average of each column
colMeansWine <- colMeans(wineDf)

# Get the average of each column per class
colMeansWineByClass <- aggregate(wineDf, by = list(wineDf$Type), FUN = mean)
colMeansWineByClass <- colMeansWineByClass[, -1]

# Bind these results into a single data frame
aveWineDf1 <- data.frame(t(colMeansWine))
aveWineDf <- rbind(colMeansWineByClass, aveWineDf1)

# Show the data frame
head(aveWineDf)
```

In this data frame we can see that rows 1, 2, and 3 represent the average of each variable for beers of type 1, 2, and 3, respectively, and the last row represents the average of each variable for all wines.

### Step 4

Now carry out a series of t-tests to examine whether the level of Ash differs across the three classes.

```{r}
# First, get 3 lists that represents the level of Ash of the 3 different kinds of wine
ash1 <- subset(wineDf$Ash, wineDf$Type == 1)
ash2 <- subset(wineDf$Ash, wineDf$Type == 2)
ash3 <- subset(wineDf$Ash, wineDf$Type == 3)

# Then, carry out the t-tests for 1~2, 2~3, 1~3
t.test(ash1, ash2)
t.test(ash2, ash3)
t.test(ash1, ash3)
```

In the t-test of these 3 pairs of independent samples, we can observe that the p-value of the t-test results of the 2 pairs of samples, 1\~2 and 2\~3, is much less than 0.05, so the null hypothesis can be rejected and it is considered that there is a significant difference between each of the 2 pairs of samples.

On the contrary, the t-test for the pair of samples 1～3 showed a p-value greater than 0.05, so the null hypothesis could not be rejected and the pair of samples could not be considered significantly different from each other.

## Problem 2

### Step 1

First, write the function.

```{r}
#' This function identifies whether a given integer is a "perfect power'
#' 
#' @param number the number to check
#' @param power the specified power
#' 
#' @return the list that shows the result
isPerfectPower <- function(number, power) {
  
  if (abs(round(number)-number != 0) >= .Machine$double.eps^0.5) {
    stop("The number must be an integer.")
  }
  if (abs(round(power)-power != 0) >= .Machine$double.eps^0.5) {
    stop("The power must be an integer.")
  }
  
  isPerfect = abs(round(number^(1/power))-number^(1/power)) < .Machine$double.eps^0.5
  root = number^(1/power)
  
  return(list(isPerfect, root))
}
```

### Step 2

Next, demonstrate our function works.

```{r}
#' This function which calls the function above, using a loop to identify both the root and power
#' 
#' @param number the input number
#' 
#' @return the root and the power as a valid string equation with an appropriate message if the input is not a perfect power
findRootPower <- function(number) {
  
  myVar <- FALSE
  powers <- c()
  
  for (i in 2:round(sqrt(number))) {
    if (isPerfectPower(number, i)[[1]] == TRUE) {
      myVar <- TRUE
      powers <- append(powers, i)
    }
  }
  
  if (myVar == TRUE) {
    selectedPower = min(powers)
    myString = paste(as.character(number), 
                     " = ",
                     as.character(number^(1/selectedPower)),
                     "^", 
                     as.character(selectedPower))
    return(myString)
  }
  else {
    return("The number entered is not a perfect power.")
  }
}
```

### Step 3

Finally, use this function to test on some examples.

```{r}
findRootPower(27)
findRootPower(13060694016)
findRootPower(7776)
findRootPower(170859375)
findRootPower(58247422)
findRootPower(94143178827)
```

## Problem 3

### Step 1

Here is the answer from GPT-4:

*Here are two R functions that accomplish the tasks you described:*

```{r}
# Function to determine the name of a poker hand
determine_poker_hand <- function(suits, ranks) {
  # Count the occurrences of each rank
  rank_counts <- table(ranks)
  
  # Check for a flush (all cards have the same suit)
  is_flush <- length(unique(suits)) == 1
  
  # Check for a straight (consecutive ranks)
  is_straight <- max(ranks) - min(ranks) == 4 && length(unique(ranks)) == 5
  
  # Determine the hand name
  if (is_flush && is_straight) {
    return("Straight Flush")  # Both straight and flush
  } else if (any(rank_counts == 4)) {
    return("Four of a Kind")  # Have 4 cards with the same kind
  } else if (all(rank_counts == c(2, 3))) {
    return("Full House")  # Have 3 cards with identical rank and 2 cards with another identical rank
  } else if (is_flush) {
    return("Flush")
  } else if (is_straight) {
    return("Straight")  
  } else if (any(rank_counts == 3)) {
    return("Three of a Kind")  # Have 3 cards with identical kind
  } else if (sum(rank_counts == 2) == 2) {
    return("Two Pair")  # Have 2 cards with identical rank and 2 cards with another identical rank
  } else if (sum(rank_counts == 2) == 1) {
    return("One Pair")  # Have only 1 pair of cards with identical rank
  } else {
    return("High Card")
  }
}

# Function to simulate dealing a round of poker
deal_poker_round <- function(num_players) {
  # Define the deck of cards
  ranks <- rep(2:14, 4) # 2 through Ace (14), each repeated 4 times
  suits <- rep(c("Hearts", "Diamonds", "Clubs", "Spades"), each = 13)
  deck <- data.frame(Rank = ranks, Suit = suits)  # Combine the suits and ranks within a single data frame
  # Shuffle the deck
  shuffled_deck <- deck[sample(nrow(deck)), ]
  
  # Deal cards to each player
  player_hands <- list()  # Initialize a empty list to store the cards of each player
  for (i in 1:num_players) {
    player_hand <- shuffled_deck[((i - 1) * 5 + 1):(i * 5), ]  # The numbers are taken sequentially from the data frame, and the process is treated as dealing cards to each players
    player_hands[[i]] <- player_hand  # append each player's cards to the list
    cat("Player", i, "hand:", "\n")  # Output the player's cards
    print(player_hand)
    hand_name <- determine_poker_hand(player_hand$Suit, player_hand$Rank)
    cat("Hand Name:", hand_name, "\n\n")  # Output the player's name of the kind
  }
}

# Example usage:
deal_poker_round(4)  # Simulate a round of poker with 4 players
```

*These functions allow you to determine the name of a poker hand based on a given set of suits and ranks, as well as simulate dealing a round of poker with a specified number of players, displaying each player's hand and its corresponding name.*

### Step 2

Test the code with a range of inputs.

```{r}
deal_poker_round(2)
deal_poker_round(3)
deal_poker_round(5)
deal_poker_round(6)
deal_poker_round(7)
```

### Step 3

I have already explained the code line by line in Step 1.

### Step 4

The function "determine_poker_hand" that ChatGPT generated above have 2 problems: a. it allows more than 10 hands to be dealt; b. it will generate the warning: "Warning: longer object length is not a multiple of shorter object length".

Fore example:

```{r}
deal_poker_round(12)
```

It shows that the 11th player is only assigned 2 cards, and that the 12th player is not assigned any cards.

To revise the function and fix the problems, I made following attempts:

```{r}
deal_poker_round_revised <- function(num_players) {
  if (num_players <= 10) {
    ranks <- rep(2:14, 4) 
    suits <- rep(c("Hearts", "Diamonds", "Clubs", "Spades"), each = 13)
    deck <- data.frame(Rank = ranks, Suit = suits)  
  
    shuffled_deck <- deck[sample(nrow(deck)), ]
  
    player_hands <- list() 
  
    for (i in 1:num_players) {
      player_hand <- shuffled_deck[((i - 1) * 5 + 1):(i * 5), ]  
      player_hands[[i]] <- player_hand 
      cat("Player", i, "hand:", "\n") 
      print(player_hand)
      hand_name <- determine_poker_hand(player_hand$Suit, player_hand$Rank)
      cat("Hand Name:", hand_name, "\n\n") 
    }
  }
  else {
    print("The number of players exceeds the limit.")
  }
}
```

Then we can check what will it happen if the number of players exceeds the limit.

```{r}
deal_poker_round_revised(11)
```

As for the warning "Warning: longer object length is not a multiple of shorter object length", the root cause of this problem is that the 'determine_poker_hand' function tries to compare a vector 'rank_counts' of indeterminate length with the vector 'c(2, 3)'. Below is how I refine the original function.

```{r}
determine_poker_hand_revised_2 <- function(suits, ranks) {
  # Count the occurrences of each rank
  rank_counts <- table(ranks)
  
  # Check for a flush (all cards have the same suit)
  is_flush <- length(unique(suits)) == 1
  
  # Check for a straight (consecutive ranks)
  is_straight <- max(ranks) - min(ranks) == 4 && length(unique(ranks)) == 5
  
  # Determine the hand name
  if (is_flush && is_straight) {
    return("Straight Flush")  # Both straight and flush
  } else if (any(rank_counts == 4)) {
    return("Four of a Kind")  # Have 4 cards with the same kind
  } else if (length(unique(rank_counts)) == 2 && all(rank_counts %in% c(2, 3))) {
    return("Full House")  # Have 3 cards with identical rank and 2 cards with another identical rank
  } else if (is_flush) {
    return("Flush")
  } else if (is_straight) {
    return("Straight")  
  } else if (any(rank_counts == 3)) {
    return("Three of a Kind")  # Have 3 cards with identical kind
  } else if (sum(rank_counts == 2) == 2) {
    return("Two Pair")  # Have 2 cards with identical rank and 2 cards with another identical rank
  } else if (sum(rank_counts == 2) == 1) {
    return("One Pair")  # Have only 1 pair of cards with identical rank
  } else {
    return("High Card")
  }
}

# Function to simulate dealing a round of poker
deal_poker_round_revised_2 <- function(num_players) {
  # Define the deck of cards
  ranks <- rep(2:14, 4) # 2 through Ace (14), each repeated 4 times
  suits <- rep(c("Hearts", "Diamonds", "Clubs", "Spades"), each = 13)
  deck <- data.frame(Rank = ranks, Suit = suits)  # Combine the suits and ranks within a single data frame
  # Shuffle the deck
  shuffled_deck <- deck[sample(nrow(deck)), ]
  
  # Deal cards to each player
  player_hands <- list()  # Initialize a empty list to store the cards of each player
  for (i in 1:num_players) {
    player_hand <- shuffled_deck[((i - 1) * 5 + 1):(i * 5), ]  # The numbers are taken sequentially from the data frame, and the process is treated as dealing cards to each players
    player_hands[[i]] <- player_hand  # append each player's cards to the list
    cat("Player", i, "hand:", "\n")  # Output the player's cards
    print(player_hand)
    hand_name <- determine_poker_hand_revised_2(player_hand$Suit, player_hand$Rank)
    cat("Hand Name:", hand_name, "\n\n")  # Output the player's name of the kind
  }
}

# Example usage:
deal_poker_round_revised_2(4)
```

As we can see, the problem is fixed now.
